{\em pyg4ometry}, a python library for the creation of Monte Carlo particle transport geometries

### 2019-02-18 paper/pyg4ometry.tex
Removed 1st year tutorial chat
Filled out all sections

\author[a]{Stewart T. Boogert\corref{author}}
\author[a]{Andrey Abramov}
\author[a]{Laurence Nevay}

\cortext[author] {Corresponding author.\\\textit{E-mail address:} stewart.boogert@rhul.ac.uk}
\address[a]{First Address}
\address[b]{Second Address}

\begin{abstract}
%% Text of abstract

\end{abstract}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
keyword1; keyword2; keyword3; etc.

@@

\begin{small}
\noindent
{\em Program Title: pyg4ometry }                                         \\
{\em Licensing provisions(please choose one): CC0 1.0/CC By 4.0/MIT/Apache-2.0/BSD 3-clause/BSD 2-clause/GPLv3/GPLv2/LGPL/CC BY NC 3.0/MPL-2.0 }                                   \\
{\em Programming language: python, C/C++}                                   \\

@@

{\em Nature of problem(approx. 50-250 words):}\\
Creating computer readable geometry descriptions for Monte Carlo particle transport codes is a time consuming and error prone task.  
{\em Solution method(approx. 50-250 words):}\\
  %Describe the method solution here.
{\em Additional comments including Restrictions and Unusual features (approx. 50-250 words):}\\
  %Provide any additional comments here.
   \\

@@

%% main text
\section{Introduction} \label{sec:introduction}
There are numerous different software codes to simulate the passage of particles through material, so called particle transport (PT) programmes, these 
include MCMP, FLUKA, Geant3 and Geant4. Each code either has a particular specialism or target user community.  A fundamental requirement of 
all of the codes is to supply a computer readable description of the physical geometry that the particles are passing through.  The creation of geometry 
files is typically a very time consuming activity, the simulation validity and performance is directly dependent on the quality of the geometry. The requirement of a 
geometry system that allows the conversion between files prepared for different codes will enable cross checks of the physics processes in particle 
transport codes. Parallel to the creation of geometry files for PT programs, usually computer assisted design (CAD) files exist for systems which 
need to be simulated in PT programmes. The fundamental geometric representations in CAD files is typically not amenable for PT programmes.  For 
these reasons it is advantageous to create a software tool that allows particle transport code users to rapidly develop geometry files, convert between 
common PT programme geometry formats and finally import CAD models. This paper describes such geometry creation and conversion package 
called {\em pyg4ometry}, implemented in {\em Python} and internally based on the Geant4 geometry description. The main features of  {\em pyg4ometry} 
are a python scripting API to rapidly design parametrised geometry, conversion to and from  FLUKA geometry descriptions, conversion from CAD formats 
(STEP and XXX) based on FreeCAD-OpenCascade and powerful geometry visualisation tools based on VTK.

@@

\section{{\em pyg4ometry} design and layout}
The core of {\em pyg4ometry} consists of python classes that wrap Geant4 solid, logical and physical volume objects.

\section{Rapid geometry modelling}

\section{CAD to Geant4 conversion }

\section{Geant4 and FLUKA conversion}

\section{Performance in Geant4}

\section{Conclusions}

>>>>

\author[a]{Stewart T. Boogert\corref{author}}
\author[a]{Andrey Abramov}
\author[a]{Laurence Nevay}
\author[a]{William Shields}

\cortext[author] {Corresponding author.\\\textit{E-mail address:} stewart.boogert@rhul.ac.uk}
\address[a]{John Adams Institute at Royal Holloway, Department of Physics, Royal Holloway, Egham, TW20 0EX, Surrey, UK}

\begin{abstract}
\end{abstract}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
Geant; Monte Carlo; Particle; Transport; Geometry;

@@

\begin{small}
\noindent
{\em Program Title: pyg4ometry }                                         \\
{\em Licensing provisions(please choose one): GPLv3 }                                   \\
{\em Programming language: python, C/C++}                                   \\

@@

{\em Nature of problem(approx. 50-250 words):}\\
Creating computer readable geometry descriptions for Monte Carlo particle transport codes is a time consuming and error prone task.  \\
{\em Solution method(approx. 50-250 words):}\\
  %Describe the method solution here.
{\em Additional comments including Restrictions and Unusual features (approx. 50-250 words):}\\
  %Provide any additional comments here.
   \\

@@

%% main text
\section{Introduction} \label{sec:introduction}
There are numerous different software codes to simulate the passage of particles through material, so called particle transport (PT) programmes, these 
include MCMP, FLUKA, Geant3 and Geant4. Each code either has a particular specialism or target user community.  PT Monte Carlo simulations have 
diverse uses, shielding calculations for radiological protection, detector performance, medical imaging and therapy, space radiation environment simulation 
are some examples. A fundamental requirement of all of the codes is to supply a computer readable description of the physical geometry that the particles 
are passing through.  The creation of geometry files is typically a very time consuming activity, the simulation validity and performance is directly dependent 
on the quality of the geometry. The requirement of a  geometry system that allows the conversion between files prepared for different codes will enable cross 
checks of the physics processes in particle transport codes. Parallel to the creation of geometry files for PT programs, usually computer assisted design (CAD) 
files exist for systems which need to be simulated in PT programmes. The fundamental geometric representations in CAD files is typically not amenable for PT 
programmes.  For these reasons it is advantageous to create a software tool that allows particle transport code users to rapidly develop geometry files, convert between 
common PT programme geometry formats and finally import CAD models. This paper describes such geometry creation and conversion package 
called {\em pyg4ometry}, implemented in {\em Python} and internally based on the Geant4 geometry description. The main features of  {\em pyg4ometry} 
are a python scripting API to rapidly design parametrised geometry, conversion to and from  FLUKA geometry descriptions, conversion from CAD formats 
(STEP and XXX) based on FreeCAD-OpenCascade and powerful geometry visualisation tools based on VTK.

@@

\subsection{Geant Detector Mark-up Language} 
To exchange geometry descriptions Geant Detector Mark-up Language (GDML) was developed. Geant4 and ROOT can read and write GDML format and it is commonly
used as an exchange format for Geant4 geometries. 
\begin{table}
\centering
\begin{tabular}{|c|c|c|} \hline
GDML tag 		&  Child tags				& pyg4ometry class 			\\ \hline
Defines 			&						&						\\
Solids 			&						&						\\
Materials 			&						&						\\
Structure			&						&						\\ \hline
\end{tabular}
\caption{GDML tags and corresponding {\it pyg4ometry} classes.}
\end{table}

\section{{\em pyg4ometry} design and layout}
The core of {\em pyg4ometry} consists of python classes that mimic Geant4 solids, logical, physical volume objects, GDML parameters material classes.
The constructors of the python classes are kept as close to the original Geant4 C++ implementation so that {\it pyg4ometry} users do not have to learn 
a new API. Geometry construction in python proceeds in a way which is very similar to geometry construction in Geant4. A user relatively familiar with Geant4
should be able to start creating geometry in pyg4ometry immediately.

\subsection{Expression engine}
In GDML symbolic expressions can be used to parametrise solids and their placement. These expressions are evaluated when the GDML is loaded into Geant4. 
In order to fully replicate the functionality of GDML an expression engine was implemented using ANTLR. GDML allows for the definition and assignment of 
variables. GDML expressions are not much more complicated than binary operators $+, -, \times, /$ and common trigonometric and special functions $\sin, \cos, \tan$ etc. 
The AST terminates on either expressions which evaluate to numbers or GDML variables. Internally all classes use GDML expressions and not floating point numbers. 
Storing internal data as expressions allows for deferred evaluation (or re-evaluation) of solid parameters and placements. This allows a user to update variables whilst
defining geometry and the expression engine will update all internal values.

\subsection{Tessellation engine}
For each Geant4 primitive solid a triangular tessellated mesh is generated and cached. This mesh is then used to determine the extent of placed instances of 
geometry (physical volumes) and meshes for CSG derived solids. CSG mesh calculations are performed using a BSP tree technique. Triangular meshes based
on CSG operations involving curved surfaces often contain large numbers of thin triangles. Before meshes are visualised or written to disk various algorithms from 
TetGEN can be employed to give the meshes more desirable output.

\subsection{Visualisation}
A  geometry hierarchy can be viewed using the popular Visualisation Toolkit (VTK). No separate scene tree is required as the Geant4 volume hierarchy is sufficient 
to place the meshes associated with each physical volume. The python version of Geant4's physical volume is also used to store visualisation attributes like the solids 
color, surface or wireframe representation and visibility.

\subsection{Internal data representation} 
The internal data representation follows closely the structure of GDML. A registry class aggregates python ordered dictionaries that are  used to store the main 
elements of a GDML file. As a user specifies geometry the registry is updated. When a user is finished with the geometry, the registry is written to disk as a GDML file.

\subsection{Input and output} 
For each input format available in {\it pyg4ometry} a dedicated {\it Reader} class is implemented, so \verb|gdml.Reader|, \verb|stl.Reader|, \verb|fluka.Reader| 
and \verb|freecad.Reader|. Each reader constructs the appropriate Geant4 classes and

\section{Rapid geometry modelling}
Given the python scripting interface, expression  and tessellation engines. It is possible for a user to rapidly specify the geometrical layout of the PT problem, vary 
the parameters of the geometry and visualise the geometry. Finally when the desired result without geometry overlaps is finalised a GDML file can be output
from the internal memory representation.

\section{CAD to Geant4 conversion }
STEP and IGES files can be loaded into pyg4ometry, via an interface based on FreeCAD. FreeCAD is an open source CAD/CAM program, which in turn 
is based on OpenCASCADE. A STEP CAD file could be considered as a hierarchical tree of {\it part assemblies} and {\it part features}, where a part assembly is a 
collection of part features.

\section{Geant4 and FLUKA conversion}
FLUKA geometry is based upon a limited set of primitive volumes which can be combined using logical operations. Many FLUKA primitives are infinite in extent, so 
to emulate FLUKA geometry construction large but finite extent Geant4 primitive solids are used. Table XXX shows the mapping between FLUKA and Geant4 objects.

\section{Performance of converted geometry in Geant4}
Clearly for Geant4 geometry defined in {\it pyg4ometry} the performance is exactly the same as it the geometry was specified in C++ and linked against the Geant4
libraries. A comparison of the performance of converted CAD or FLUKA geometry compared with native Geant4 solids is required so a user can optimise the available
computing resources against the human resources (required to convert geometry).

\section{Conclusions}
The authors believe that tools to quickly create, either ab-initio or by conversion geometry usable in particle transport Monte Carlo programmes will save significant 
amounts of user effort and ultimately yield more accurate simulations. {\it pyg4ometry} is a relatively complete implementation of a geometry creation tool, whilst heavily 
internally based on Geant4 and GDML can have utility for users of all PT codes. {\it pyg4ometry} can clearly be extended to other formats or applications. As for applications,
examples include  the  generation triangular mesh structures for GPU accelerated photon tracking in liquid noble  dark matter detectors and geometries for event visualisation.
